---
title       : SVD - Singular Value Decomposition
subtitle    : Practical Intro
author      : Mykola Pavlov
job         : Biomech, Inc
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]     # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## What is SVD?

$$latex
A = U \Sigma V^{T}
$$

$A$ is m x n matrix

$U$ is m x k orthonormal matrix 

$V$ is n x k orthonormal matrix

$\Sigma$ is k x k - diagonal matrix

---

## Latent Semantic Analysis

Application of SVD to term-document matrices

``` {r, echo=FALSE}

docs <- c("Human machine interface for Lab ABC computer applications",
        "A survey of user opinion of computer system response time",
        "The EPS user interface management system",
        "System and human system engineering testing of EPS",
        "Relation of user-perceived response time to error measurement",
        "The generation of random, binary, unordered trees",
        "The intersection graph of paths in trees",
        "Graph minors IV: Widths of trees and well-quasi-ordering",
        "Graph minors: A survey")
docs

```

Documents (Deerwester et al. (1990), Table 2, Part 1)

---

## Term-Document Matrix

``` {r, cache=TRUE}
df <- read.csv(file="matrix.csv")
print(df, row.names=F)
```

---

## Let's compute SVD

``` {r, echo=FALSE}
options(digits=1)
```

``` {r}
df.svd <- svd(df[,-1])
scales <- df.svd$d[1:2]
scales

``` 

--- 

## Term Space

``` {r}
termVectors <- df.svd$u[,1:2]
terms.lsi <- data.frame(termVectors, terms)
print(terms.lsi, row.names=F)
```

---

## Document Space

``` {r}
docVectors <- df.svd$v[,1:2]
docs.lsi <- data.frame(docVectors, Docs=strtrim(docs, 50))
print(docs.lsi, row.names=F, col.names=F)
```

---

## Query "Human computer interaction"

``` {r}
human <- as.numeric(terms.lsi[terms.lsi$terms == "human", 1:2])
computer <- as.numeric(terms.lsi[terms.lsi$terms == "computer", 1:2])
query <- human + computer
print(query)
```

$(0.22, -0.11) + (0.24, 0.04) = (0.46, -0.07)$

---

## Document Scores VS Query

Cosine distance: $Cos\theta =  \frac{a \cdot b}{||a|| ||b||}$

``` {r}
y <- (scales * query)
scores <- apply(docs.lsi[,1:2], 1, function(x) x %*% y / sqrt(x%*%x * y%*%y))
print(data.frame(scores, docs), row.names=F)
```

--- 

## Query and Document Space

``` {r}
plot(docs.lsi$X1, docs.lsi$X2, type="n", xlim = c(-1.6, 0))
arrows(0,0, docs.lsi$X1, docs.lsi$X2)
arrows(0,0, scales[1]*query[1], scales[2]*query[2], col=2)
```

---

## Term Scores VS Query

``` {r}
y <- (scales * query)
scores <- apply(terms.lsi[,1:2], 1, function(x) x %*% y / sqrt(x%*%x * y%*%y))
print(data.frame(scores, terms), row.names=F)
```

---

## Let's dive into math
Row Space -> Column Space

$AV = U \Sigma$

``` {r, echo=FALSE}
print(df, row.names=F)
```

---

## Simplifications 

$A = U \Sigma V^{-1} = U \Sigma V^{T}$
  
$A^{T} A = V \Sigma^{T} U^{T} U \Sigma V^{T} = V \Sigma^2 V^{T}$

$A A^{T} = U \Sigma V^{T} V \Sigma^{T} U^{T}= U \Sigma^2 U^{T}$

But what is $U$, $\Sigma$, $V$?

--- 

## Matrix Spectral Theorem

There exists an orthonormal basis of Q consisting of eigenvectors of B. Each eigenvalue is real

$B = Q \Lambda Q^{T}$

And our $AA^{T}$ is like B, squared, symmetric 

But what is eigenvalues and eigenvectors?

--- 

## Eigenvalues

$Ax = \lambda x$ 

<img class=center src=assets/fig/eigen.gif height='400'/>

--- 

## Solve 

$Ax - \lambda x = 0$

$(A - \lambda I) x = 0$

$A - \lambda I = 0$

$\lambda$ - eigenvalues

$x$ - eigenvectors

--- 

## Algorithm to find SVD

 1. Find eigenvectors $U$ of $A A^{T}$ 
 
 2. Find eigenvectors $V$ of $A^{T} A$
 
 3. The non-zero singular values of $A$ found on the diagonal entries of $\Sigma$ are the square roots of the non-zero eigenvalues of both $A A^{T}$ and $A^{T} A$.

That's it!

---  
 
## Eigenvalues

<img class=center src=assets/fig/eigenval.png height='512'/>

---
## Links

MIT Professor Gilbert Strang teaching 18.06:
http://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/video-lectures/

SVD Tutorial:
http://alias-i.com/lingpipe/demos/tutorial/svd/read-me.html

Wiki: http://en.wikipedia.org/wiki/Singular_value_decomposition

---
## Contacts

### Nikolay Pavlov, Founder at Biomech
 
* E-mail: me@nikolaypavlov.com 
* Linkedin: http://linkedin.com/in/nikolaypavlov
* Facebook: http://www.facebook.com/pav.nikolay
* Twitter: https://twitter.com/nikolaypavlov
* Github: https://github.com/nikolaypavlov

